Moment Distances from Robust Subspace for Botnet Detection

### 2019-05-15 docs/papers/2019JNCA_m-rpca/2019JNCA_m-rpca.tex
working on results

Mahalanobis Distance (MD) is a generalized distance which is useful for determining the similarity between an unknown sample and a collection of known samples, by considering the correlations between the variables and their mean values. The MD is a measure of the distance between a vector $\boldsymbol{x}$ and a distribution $\boldsymbol{X}$, introduced by P. C. Mahalanobis in 1936 \cite{mahalanobis1936md}. It is a multi-dimensional generalization for measuring how many standard deviations away $\boldsymbol{x}$ is from the mean $\boldsymbol{\bar{x}}$ and covariance $\boldsymbol{\bar{\Sigma}}$ of $\boldsymbol{X}$. MD has been used for distance based anomaly detection with robust estimates in many areas, assuming that the Mahalanobis Distance between robust estimates and new observations can reveal anomalies.

@@

Robust PCA is a well known method to recover a low-rank matrix $\textbf{L}$ and sparce matrix $\textbf{S}$ from corrupted measurements modeled as $\textbf{X} = \textbf{L} + \textbf{S}$. This decomposition in low-rank and sparse matrices can be achieved by techniques such as Principal Component Pursuit method (PCP), and by optimization methods, such as the Augmented Lagrange Multiplier Method (ALM), Alternating Direction Method (ADM), Fast Alternating Minimization (FAM) or Iteratively Reweighted Least Squares (IRLS) \cite{candes2011robust,vaswani2018robust,lerman2018overview}. We adopt RPCA with ALM, which can be formulated as

\begin{equation}\label{eq:eq01}
	\min_{\boldsymbol{L,S}}\left\|\boldsymbol{L}\right\|_* + \lambda\left\|\boldsymbol{S}\right\|_1 + \langle \boldsymbol{Y, X - L - S}  \rangle + \frac{\mu}{2}\left\|\boldsymbol{X - L - S}\right\|_F^2,
\end{equation}
where $\boldsymbol{Y}$ are the Lagrange multipliers and have the constraint $\boldsymbol{C = D - A - E}$...

@@

The general formula of the $u$-th moment can be expressed as
\begin{equation}\label{eq:eq04}
	m_u = \displaystyle\frac{1}{n}\displaystyle\sum_{i = 1}^{n}( x_i - \bar{x})^u.
\end{equation}

Therefore, the skewness is calculated by
\begin{equation}\label{eq:eq05}
	\boldsymbol{\bar{s}} = \frac{m_3}{m_2^{\frac{3}{2}}},
\end{equation}

and the kurtosis as
\begin{equation}\label{eq:eq06}
	\boldsymbol{\bar{k}} = \frac{m_4}{m_2^2}.
\end{equation}

@@

The classical Mahalanobis Distance is defined as		
\begin{equation}\label{eq:eq07}
	\boldsymbol{d}(\boldsymbol{x},\bar{\boldsymbol{x}}, \boldsymbol{\bar{\Sigma}}) = \sqrt{(\boldsymbol{x} - \bar{\boldsymbol{x}}) \boldsymbol{\bar{\Sigma}}^{-1}(\boldsymbol{x} - \bar{\boldsymbol{x}})^\prime},
\end{equation}
where $\boldsymbol{x}$ is a vector of a new observation, $\bar{\boldsymbol{x}}$ is the mean vector of known observations, also referred as location, and $\boldsymbol{\bar{\Sigma}}$ is the covariance matrix of known observations, also referred as scatter. The classical MD usually relies on robust mean and robust covariance matrix for outlier detection, which are commonly computed by Fast-MCD \cite{rousseeuw1984mcd, rousseeuw1999fastmcd}. Here we propose to calculate the MD based on mean and covariance matrix computed from a robust subspace learned by RPCA. We also propose to extend the Equation \ref{eq:eq05} to implement a Skewness-based Mahalanobis Distance, as follows:

@@

The contamination rate parameter is widely adopted for well established outlier detection algorithms \cite{zhao2019pyod}, and refers to the percentage of observations that are classified as anomalous. The contamination rate can be well known for some areas, or can be computed by cross-validation or can be assumed according to previous observations. In our proposal, the contamination define the rate of highest distances $\boldsymbol{d}$ that shall be classified as anomalous. Therefore, the resulting $\boldsymbol{\bar{t}}$ refers to observations classified as anomalous according to $c$.

The Algorithm \ref{alg:alg01} presents how the proposed Moment Distances from Robust Subspace for Botnet Detection can be algorithmically described.
\begin{algorithm}
	\label{alg:alg01}
	\SetAlgoLined
	\KwResult{$\boldsymbol{\bar{t}}$}
	Given $\boldsymbol{X}$ split into $\boldsymbol{X}^n$ and $\boldsymbol{X}_i^t$\;
	$\min_{\boldsymbol{L}^n, \boldsymbol{S}^n}\left\|\boldsymbol{L}^n\right\|_* + \lambda\left\|\boldsymbol{S}^n\right\|_1 + \langle \boldsymbol{Y}^n, \boldsymbol{X}^n - \boldsymbol{L}^n - \boldsymbol{S}^n  \rangle + \frac{\mu}{2}\left\|\boldsymbol{X}^n - \boldsymbol{L}^n - \boldsymbol{S}^n\right\|_F^2$\;
	$\bar{\boldsymbol{x}} = \displaystyle\frac{1}{h}\displaystyle\sum_{i\in \boldsymbol{h}} \boldsymbol{x}_i$, for $\boldsymbol{L}^n$\;
	$\boldsymbol{\bar{\Sigma}} = \displaystyle\frac{1}{h}\displaystyle\sum_{i\in \boldsymbol{h}} (\boldsymbol{x}_i - \bar{\boldsymbol{x}})(\boldsymbol{x}_i - \bar{\boldsymbol{x}})^\prime$, for $\boldsymbol{L}^n$\;
	$\boldsymbol{\bar{s}} = \frac{m_3}{m_2^{\frac{3}{2}}}$, for $\boldsymbol{L}^n$\;
	$\boldsymbol{\bar{k}} = \frac{m_4}{m_2^2}$, for $\boldsymbol{L}^n$\;
	$\boldsymbol{d}(\boldsymbol{x}, \bar{\boldsymbol{x}}, \boldsymbol{\bar{\Sigma}}) = \sqrt{(\boldsymbol{x} - \bar{\boldsymbol{x}}) \boldsymbol{\bar{\Sigma}}^{-1}(\boldsymbol{x} - \bar{\boldsymbol{x}})^\prime}$\;
	$\boldsymbol{d}(\boldsymbol{x}, \bar{\boldsymbol{s}}, \boldsymbol{\bar{\Sigma}}) = \sqrt{(\boldsymbol{x} - \bar{\boldsymbol{s}}) \boldsymbol{\bar{\Sigma}}^{-1}(\boldsymbol{x} - \bar{\boldsymbol{s}})^\prime}$\;
	$\boldsymbol{d}(\boldsymbol{x}, \bar{\boldsymbol{k}}, \boldsymbol{\bar{\Sigma}}) = \sqrt{(\boldsymbol{x} - \bar{\boldsymbol{k}}) \boldsymbol{\bar{\Sigma}}^{-1}(\boldsymbol{x} - \bar{\boldsymbol{k}})^\prime}$\;
	$\boldsymbol{\bar{t}}_{\bar{\boldsymbol{x}}} = [\boldsymbol{d}(\boldsymbol{x}, \bar{\boldsymbol{x}}, \boldsymbol{\bar{\Sigma}})]^c$\;
	$\boldsymbol{\bar{t}}_{\bar{\boldsymbol{s}}} = [\boldsymbol{d}(\boldsymbol{x}, \bar{\boldsymbol{s}}, \boldsymbol{\bar{\Sigma}})]^c$\;
	$\boldsymbol{\bar{t}}_{\bar{\boldsymbol{k}}} = [\boldsymbol{d}(\boldsymbol{x}, \bar{\boldsymbol{k}}, \boldsymbol{\bar{\Sigma}})]^c$\;
	\caption{Moment Distances from Robust Subspace}
\end{algorithm}

>>>>

Mahalanobis Distance (MD) is a generalized distance which is useful for determining the similarity between an unknown sample and a collection of known samples, by considering the covariances between the variables and their mean values. The MD is a measure of the distance between a vector $\boldsymbol{x}$ and a distribution $\boldsymbol{X}$, introduced by P. C. Mahalanobis in 1936 \cite{mahalanobis1936md}. It is a multi-dimensional generalization for measuring how many standard deviations away $\boldsymbol{x}$ is from the mean $\boldsymbol{\bar{x}}$ and covariance $\boldsymbol{\bar{\Sigma}}$ of $\boldsymbol{X}$. MD has been used for distance based anomaly detection with robust estimates in many areas, assuming that the Mahalanobis Distance between robust estimates and new observations can reveal anomalies.

@@

Robust PCA is a well known method to recover a low-rank matrix $\textbf{L}$ and sparse matrix $\textbf{S}$ from corrupted measurements modeled as $\textbf{X} = \textbf{L} + \textbf{S}$. This decomposition in low-rank and sparse matrices can be achieved by techniques such as Principal Component Pursuit method (PCP), and by optimization methods, such as the Augmented Lagrange Multiplier Method (ALM), Alternating Direction Method (ADM), Fast Alternating Minimization (FAM) or Iteratively Reweighted Least Squares (IRLS) \cite{candes2011robust,vaswani2018robust,lerman2018overview}.

According to Wright et al. \cite{rpca_convex_opt}, under rather broad conditions, as long as the error matrix S is sufficiently sparse, it is possible to recover a low-rank matrix by solving the following convex optimization problem:

\begin{equation}\label{eq:pcp}
	(\boldsymbol{\hat{L}}, \boldsymbol{\hat{S}})\leftarrow min_{L,S}\left \| \boldsymbol{L} \right \|_{*} + \lambda \left \| \boldsymbol{S} \right \|_{1}
\end{equation}
\begin{center} subject to: $\boldsymbol{X} = \boldsymbol{L} + \boldsymbol{S}$ \end{center}

\begin{equation}\label{eq:nuclear_norm}
    \left \| \boldsymbol{L} \right \|_{*} = \sum_{i} \sigma_{i}(\boldsymbol{L})
\end{equation}

\begin{equation}\label{eq:l1_norm}
    \left \| \boldsymbol{S} \right \|_{1} = \sum_{ij} \left | \boldsymbol{S}_{ij} \right |
\end{equation}
where $\lambda$ a positive weighting parameter, which determines the sparsity of $\boldsymbol{S}$.

 Before ALM, some methods were proposed to solve that convex optimization problem, such as Iterative Thresholding (IT) and Accelerated Proximal Gradient (APG). However, according to \cite{alm} both have scalability problems and require a large number of iterations to converge. The Augmented Lagrange Multiplier is proven to have a \emph{Q-linear} convergence speed and experimental results show that ALM is five times faster than APG, which in theory is sub-linear \cite{alm}. Furthermore, ALM reaches more accurate results with less iterations.

% We adopt RPCA with ALM, which can be formulated as

Augmented Lagrange Multiplier formula is given by:
\begin{equation}\label{eq:eq01}
	l(\boldsymbol{L}, \boldsymbol{S}, \boldsymbol{Y}) = \left\|\boldsymbol{L}\right\|_* + \lambda\left\|\boldsymbol{S}\right\|_1 + \langle \boldsymbol{Y, X - L - S}  \rangle + \frac{\mu}{2}\left\|\boldsymbol{X - L - S}\right\|_F^2,
\end{equation}
%where $\boldsymbol{Y}$ are the Lagrange multipliers and have the constraint %$\boldsymbol{C = D - A - E}$...

where

\begin{equation}\label{eq:y_cond}
	\left\{\begin{matrix} \boldsymbol{Y}_{k+1} = \boldsymbol{Y}_{k} + \mu(\boldsymbol{X} - \boldsymbol{L}_{k} - \boldsymbol{S}_{k}) \\ \boldsymbol{Y}_{k+1} \in argmin_{\boldsymbol{L,S} \in R^{m \times n}} \{l(\boldsymbol{L}, \boldsymbol{S}, \boldsymbol{Y}_{k})\} \end{matrix}\right. 
\end{equation}
$\boldsymbol{Y}$ the multiplier of the linear constraint and $\mu$ the parameter for the violation of the linear constraint \cite{adm}.

We adopt RPCA with ADM, which, generally speaking, is a practical improvement of the classical Augmented Lagrangian method for solving convex programming problem with linear constraints, by fully taking advantage of its high-level separable structure \cite{adm}. ADM minimizes $\boldsymbol{L}$ and $\boldsymbol{S}$ variables serially by solving the following problems to generate the new iterate:

\begin{equation}\label{eq:adm_eq}
	\left\{\begin{matrix}
	\boldsymbol{L}_{k+1} \in argmin_{\boldsymbol{L} \in R^{m \times n}}\{l(\boldsymbol{L}, \boldsymbol{S}_{k}, \boldsymbol{Y}_{k})\}\\ 
	\boldsymbol{S}_{k+1} \in argmin_{\boldsymbol{S} \in R^{m \times n}}\{l(\boldsymbol{L}_{k+1}, \boldsymbol{S}, \boldsymbol{Y}_{k})\}\\ 
	\boldsymbol{Y}_{k+1} = \boldsymbol{Y}_{k} + \mu(\boldsymbol{X} - \boldsymbol{L}_{k} - \boldsymbol{S}_{k})
	\end{matrix}\right.
\end{equation}

@@

The general formula of the $u$-th moment can be expressed as
\begin{equation}\label{eq:eq02}
	m_u = \displaystyle\frac{1}{n}\displaystyle\sum_{i = 1}^{n}( x_i - \bar{x})^u.
\end{equation}

Therefore, the skewness is calculated by
\begin{equation}\label{eq:eq03}
	\boldsymbol{\bar{s}} = \frac{m_3}{m_2^{\frac{3}{2}}},
\end{equation}

and the kurtosis as
\begin{equation}\label{eq:eq04}
	\boldsymbol{\bar{k}} = \frac{m_4}{m_2^2}.
\end{equation}

@@

The classical Mahalanobis Distance is defined as		
\begin{equation}\label{eq:eq05}
	\boldsymbol{d}(\boldsymbol{x},\bar{\boldsymbol{x}}, \boldsymbol{\bar{\Sigma}}) = \sqrt{(\boldsymbol{x} - \bar{\boldsymbol{x}}) \boldsymbol{\bar{\Sigma}}^{-1}(\boldsymbol{x} - \bar{\boldsymbol{x}})^\prime},
\end{equation}
where $\boldsymbol{x}$ is a vector of a new observation, $\bar{\boldsymbol{x}}$ is the mean vector of known observations, also referred as location, and $\boldsymbol{\bar{\Sigma}}$ is the covariance matrix of known observations, also referred as scatter. The classical MD usually relies on robust mean and robust covariance matrix for outlier detection, which are commonly computed by Fast-MCD \cite{rousseeuw1984mcd, rousseeuw1999fastmcd}. Here we propose to calculate the MD based on mean and covariance matrix computed from a robust subspace learned by RPCA. We also propose to extend the Equation \ref{eq:eq05} to implement a Skewness-based Mahalanobis Distance, as follows:

@@

The contamination rate parameter is widely adopted for well established outlier detection algorithms \cite{zhao2019pyod}, and refers to the percentage of observations that are classified as anomalous. The contamination rate can be well known for some areas, or can be computed by cross-validation or can be assumed according to previous observations. In our proposal, the contamination define the rate of highest distances $\boldsymbol{d}$ that shall be classified as anomalous. Finally, the resulting $\boldsymbol{\bar{t}}$ refers to the estimated classification of new observations, as normal or anomalous, according to $c$.

The Algorithm \ref{alg:alg01} presents how the proposed Moment Distances from Robust Subspace for Botnet Detection can be algorithmically described.
\begin{algorithm}
	\label{alg:alg01}
	\scriptsize
	\SetAlgoLined
	\KwResult{$\boldsymbol{\bar{t}}$}
	Given $\boldsymbol{X}$ split into $\boldsymbol{X}^n$ and $\boldsymbol{X}_i^t$\;
	$\min_{\boldsymbol{L}^n, \boldsymbol{S}^n}\left\|\boldsymbol{L}^n\right\|_* + \lambda\left\|\boldsymbol{S}^n\right\|_1 + \langle \boldsymbol{Y}^n, \boldsymbol{X}^n - \boldsymbol{L}^n - \boldsymbol{S}^n  \rangle + \frac{\mu}{2}\left\|\boldsymbol{X}^n - \boldsymbol{L}^n - \boldsymbol{S}^n\right\|_F^2$\;
	$\bar{\boldsymbol{x}} = \displaystyle\frac{1}{h}\displaystyle\sum_{i\in \boldsymbol{h}} \boldsymbol{x}_i$, for $\boldsymbol{L}^n$\;
	$\boldsymbol{\bar{\Sigma}} = \displaystyle\frac{1}{h}\displaystyle\sum_{i\in \boldsymbol{h}} (\boldsymbol{x}_i - \bar{\boldsymbol{x}})(\boldsymbol{x}_i - \bar{\boldsymbol{x}})^\prime$, for $\boldsymbol{L}^n$\;
	$\boldsymbol{\bar{s}} = \frac{m_3}{m_2^{\frac{3}{2}}}$, for $\boldsymbol{L}^n$\;
	$\boldsymbol{\bar{k}} = \frac{m_4}{m_2^2}$, for $\boldsymbol{L}^n$\;

	$\boldsymbol{d}(\boldsymbol{x},\bar{\boldsymbol{x}}, \boldsymbol{\bar{\Sigma}}) = \sqrt{(\boldsymbol{x} - \bar{\boldsymbol{x}}) \boldsymbol{\bar{\Sigma}}^{-1}(\boldsymbol{x} - \bar{\boldsymbol{x}})^\prime}$\;
	$\boldsymbol{d}(\boldsymbol{x}, \bar{\boldsymbol{s}}, \boldsymbol{\bar{\Sigma}}) = \sqrt{(\boldsymbol{x} - \bar{\boldsymbol{s}}) \boldsymbol{\bar{\Sigma}}^{-1}(\boldsymbol{x} - \bar{\boldsymbol{s}})^\prime}$\;
	$\boldsymbol{d}(\boldsymbol{x}, \bar{\boldsymbol{k}}, \boldsymbol{\bar{\Sigma}}) = \sqrt{(\boldsymbol{x} - \bar{\boldsymbol{k}}) \boldsymbol{\bar{\Sigma}}^{-1}(\boldsymbol{x} - \bar{\boldsymbol{k}})^\prime}$\;

	% Sort $\boldsymbol{d}_1( i )$, which yields an index $\boldsymbol{d}_1(\pi(1)) \leq ... \leq \boldsymbol{d}_1(\pi(n))$\;
	% Put $\boldsymbol{h}_2 = \{ \pi(1), \pi(2), ..., \pi(h)\}$\;
	% Compute $\boldsymbol{\bar{x}}_2$ and $\boldsymbol{\hat{S}}_2$ for $\boldsymbol{h}_2$\;
	% Compute $det(\boldsymbol{\hat{S}}_2)$ and $det(\boldsymbol{\hat{S}}_1)$\;	
	% Compute $\boldsymbol{\bar{s}}_2$ and $\boldsymbol{\bar{k}}_2$ for last $\boldsymbol{h}_2$\;
	\caption{Moment Distances from Robust Subspace for Botnet Detection}
\end{algorithm}
